Data Management and Domain Adjusting Strategies for Implementing Parallel SPH Method in JPUE Simulation

### 2016-01-07 Paper for submission.tex
added plume figure

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Zhixuan Cao, Abani Patra, Matt Jones, ...}
\IEEEauthorblockA{Department of Mechanical and Aerospace Engineering\\
 The University at Buffalo\\
 Buffalo, New York 14260\\
Email: zhixuanc@buffalo.edu, abani@buffalo.edu}
%\and
%\IEEEauthorblockN{Homer Simpson}
%\IEEEauthorblockA{Twentieth Century Fox\\
%Springfield, USA\\
%Email: homer@thesimpsons.com}
%\and
%\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
%\IEEEauthorblockA{Starfleet Academy\\
%San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212\\
%Fax: (888) 555--1212}
}
% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
This paper presents a parallel implementation of smoothed particle hydrodynamics (SPH) method using the message passing interface (MPI) standard to simulate a turbulent jet or plume which is ejected from a nozzle into a uniform environment(JPUE). Background grid is used to reduce neighbors searching cost and decompose (redecompose) the domain. Space Filing Curve(SFC) based index is adopted to assign an unique identifier to each background grid. As simulation of JUPE requires adding of new particles during simulation, time-dependent SFC based indexes are assigned to particles to guarantee uniqueness of identifiers.  
%to distinguish particles added at the same position but different simulation time step. 
Both particle and background grid are managed by hashtables which ensure quick and flexible accessing, adding and deleting of element. %Additional link list attached to the hashtable is used to handle hash conflicts.
 A SFC based three dimensional domain decomposition and a dynamic load balancing strategy are implemented to ensure good load balance. In addition, computational domain is adjusted during simulation to reduce computational cost. 
Numerical tests show that our code has acceptable strong scalability and good weak scalability.
These strategies can be further applied to many other implementations of meshfree methods, especially those implementations that require irregular
particle distributions, adding and deleting of particles and leverage adaptive background grids.
\end{abstract}

@@

SPH (spherical particle hydrodynamics) is a meshless scheme, initially developed for astrophysical applications by Lucy\cite{lucy1977numerical} and Gingold and Monaghan\cite{gingold1977smoothed}. 
Subsequently it was extended to large strain solid mechanics and computational fluid mechanics. 

@@

 simple scenarios like breaking-waves, floodings etc. Work on more complicated problems, such as eruption of volcano plume which is essentially a mult-phase turbulent flow,

@@

  Prediction of such  c complicated phenomena with acceptable accuracy at given time window requires resolutions (very high particle counts) that cannot been accomplished without parallel computing. What's more, imposing of some types of boundary conditions (such as realistic wind field, eruption boundary condition) requires dynamically adding and removing of particles during simulation. This requires efficient and more flexible data management scheme. What we need to point out is that the benefit of flexible data management strategies are not only limited to specific implementations of SPH. Actually, flexibility in data management are more critical for several advanced techniques of SPH, such as dynamic particle splitting techniques\cite{vacondio2012accurate, feldman2007dynamic}, which will give greater resolution at the area of interest by splitting one large particle to several smaller ones.

@@

To the best of the author's knowledge, no implementation of SPH has the feature of adjusting computational domain based on simulation needs. For JPUE simulation, such feature will greatly reduce computational cost by avoiding computing of uninfluenced fluid. This feature is accomplished by adding a scan function to monitor the outermost layer of the domain and turn ghost particles to real particles at the proper time. 

@@

The data structure, particle and bucket indexing strategies, domain decomposition , dynamic load balancing method and domain adjusting strategies in this paper can be easily adopted by other implementations of any meshfree methods(include SPH). The flexibility of data accessing enables implementing of meshless methods for solving of more complicated problems and using of more advanced techniques.\\

@@

\caption{Basic work flow for SPH}

@@

SPH is a meshfree, Lagrangian method. The domain is discretized by particles and the position of each particle is updated at every time step. The physical laws (such as conservation laws of mass, momentum and energy) written in the form of partial differential equations need to be transformed into the Lagrangian particle formalism of SPH. Using a kernel function that provides the weighted estimate of the field variables in the neighborhood of a discrete point (particle), the integral equations are evaluated as sums over neighbor particles. Only particles located within support of kernel function will interact. Thus, physical properties (position, density, velocity, internal energy, pressure) are updated based on its neighbors. So a neighbor search needs to be carried out before updating of physical properties. We use buckets which contain all particles associated with a sub-domain 

@@

SFC-based indexing scheme can guarantee uniqueness of particle identifier only in simple scenarios when particles are added once while setting up initial condition. In some situations, new particles need to be added while simulation. For example, new particles need to be added at the bottom of the eject vent for JPUE simulation. To distinguish particles added at the ``same place" (the small area, all points within which will be mapped to the same $k$.) at different time steps, we extend the SFC-based key to time-dependent SFC based key by including date of birth of particles into the key. The time-dependent SFC based key can be written as: $[k,t]$, where $t$ is the time step. The map $h_n$ will become:

@@

We need to emphasize here is that in theory motion of particles will destroy locality established based on initial coordinates of particles. However, as particles are moving fairly regularly in a JPUE, the locality of most of particles is largely conserved during simulation. We will revisit this question for more chaotic flows and 

@@

The most basic data structure of SPH are particle, for problem description, and bucket, for neighbors searches and domain decomposition. Both are defined as classes in C++. Information that is contained in particle class can be categorized into six categories: ID(the key), affiliate(rank of the process that the particle belongs to), primitive variables (variables show up as unknowns in governing equations, e.g. density, velocity, energy), secondary variables (properties that can be computed from primitive variables, they are stored to avoid repeated computations eg. pressure, temperature), flags (indicators, such as indicator for ghost particle and real particles, indicator for particles of different phases  ) and neighbor information (it is a vector of particle keys in our application). Similarly,  Information that is contained in  a bucket class can also be categorized into different categories: ID(the key), affiliate(rank of the process that the bucket belong to), domain information (maximum and minimum coordinates, boundary information), flags (indicators, such as indicator for guest and non-guest, indicator for active and inactive), neighbor information (keys of 27 neighbor buckets for three dimension and keys of 9 neighbor buckets for two dimension including its own key) and owned particles. Objects defined based on these two classes are then accessed through hash tables.

@@

One way to handle hash conflicts is by using an additional sorted vector attached to the hash table. When several keys hash to the same slot, a vector will be created. The vector is sorted based on keys so that a binary search can be used to find the correct position for adding, deleting or retrieving. Another option to handle hash conflicts is using an additional link list which is more flexible in memory allocation. The average time complexity of binary search is O(log n) while that for linear search based on link list is O(n). However, accessing efficiency of link list is much lower than array based data structure, especially when the link list becomes longer. Choosing of proper way to handle hash conflicts greatly depends on the problem itself. For the test problem in this paper, successively adding particles at the bottom of the eject vent will lead to hash conflicts of many particles, which implies a long link list. But considering the very long conflicts only occur on several slots among millions (see figure \ref{fig:Particle_adding_with_link}), we still choose the link list to handle hash conflicts. This decision was made based on numerical experiments.\\

@@

Particles used in the test problem can be categorized into four types based on the particle-type-flag: real particle, wall ghost particle, pressure ghost particle and eruption ghost particle. Ghost particles are for imposing of corresponding boundary conditions (see figure \ref{fig:Domain_adjusting}). As different types of particles involve different amount of computational work, shown by table \ref{tab:Computational_cost_particles} and table \ref{tab:Computational_cost_steps}, we assign different work load weight for different types of particles based on profilling data. Instead of simply using number of contained particles as work load for bucket, work load of each bucket is determined by summing up work load weight of all particles within the bucket. The SFC sequence passing through centroids of all buckets now becomes a weighted sequence. Domain decomposition  is conducted based on the weighted SFC of buckets.\\

@@

    Particle type & NS & UPME & UPD & UPP &VF &WBC &Total\\

@@

    wall ghost & No & No & No & No & No & Yes &0.75\\

@@

    eruption ghost & No & No & No & Yes & No & No & 0.02\\

@@

    pressure ghost & No & No & No & No & No & No & 0.00\\

@@

As some of the neighbor particles reside in other
partitions. A set of  particles and buckets are used to synchronize data across partitions. To minimize communications, data is synchronized only where needed, using non-blocking MPI communications. 

@@

@@

\includegraphics[scale=0.33]{Work_flow_adjust}

@@

\subfloat[Computational domain at $t=t0$]{\includegraphics[scale=0.24]{Involved_initial}}
\subfloat[Computational domain at $t=t0 + \delta t$]{\includegraphics[scale = 0.24]{Involved_domain_adjusting}}

@@

The work flow with domain adjusting is in figure \ref{fig:Work_flow_adjust}\\

@@

In current model the initial domain is a 3D box. The boundaries are categorized into eruption vent (a circle area at the center of the bottom), wall boundary (box bottom), pressure boundary (Other faces of the box).
At the vent, $T$, $\textbf{v}=\{0,0,150\}^T$, $p=1.01\times10^5 Pa$, $n_{g0}=0.05$ and mass discharge rate $\dot M$ is given. The radius of vent is determined from $\rho$, $\dot M$ and $\textbf{v}$.
Velocity is zero for non-slip wall boundary. We assume the boundary to be adiabatic and the heat flux is zero on the boundary. The pressure of the surrounding atmosphere is specified on pressure boundaries. Except for the pressure, density, velocity, and energy will depend on the solution. We adpot ghost particles to impose these different type of boundary conditions (see figure \ref{fig:Domain_adjusting}). 

@@

There are several review papers \cite{monaghan1992smoothed, monaghan2005smoothed, price2012smoothed, rosswog2009astrophysical, monaghan2012smoothed} which provide a pretty comprehensive descriptions of SPH.  These  may be consulted for additional detail.%We will not cover the basic theory of SPH in this paper.

@@

\textbf{v}_{\alpha b}=\textbf{v}_{\alpha}-\textbf{v}_b
\textbf{v}_{\alpha j}=\textbf{v}_{\alpha}-\textbf{v}_j

@@

\dfrac{d \textbf{r}_a}{dt} = \textbf{v} \label{eq:gov-update-pos}

@@

>>>>

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Zhixuan Cao, Abani Patra}
\IEEEauthorblockA{Department of MAE \\%Mechanical and Aerospace Engineering
 The University at Buffalo\\
 Buffalo, New York 14260\\
Email: zhixuanc@buffalo.edu, abani@buffalo.edu}
\and
\IEEEauthorblockN{Matthew Jones, Jaroslaw Zola}
\IEEEauthorblockA{Center for Computational Research\\
Department of CSE \\%Computer Science and Engineering
The University at Buffalo\\
Buffalo, New York 14260\\
Email: jonesm@buffalo.edu, \\ jzola@buffalo.edu}
%\and
%\IEEEauthorblockN{Jaroslaw Zola}
%\IEEEauthorblockA{The University at Buffalo\\
%Department of Computer Science and Engineering\\
% The University at Buffalo\\
% Buffalo, New York 14260\\
%Email:jzola@buffalo.edu}
}
% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
This paper presents a parallel implementation of smoothed particle hydrodynamics (SPH) method using the message passing interface (MPI) standard to simulate a turbulent jet or plume which is ejected from a nozzle into a uniform environment(JPUE). Background grid is used to reduce neighbors searching cost and to decompose (redecompose) the domain. Space Filing Curve(SFC) based index is adopted to assign an unique identifier to each background grid. As simulation of JUPE requires adding of new particles during simulation, time-dependent SFC based indexes are assigned to particles to guarantee uniqueness of identifiers.  
%to distinguish particles added at the same position but different simulation time step. 
Both particle and background grid are managed by hashtables which ensure quick and flexible accessing, adding and deleting of element. %Additional link list attached to the hashtable is used to handle hash conflicts.
A SFC based three dimensional domain decomposition and a dynamic load balancing strategy are implemented to ensure good load balance. In addition, computational domain is adjusted during simulation to reduce computational cost. 
Numerical tests show that our code has acceptable strong scalability and good weak scalability.
These strategies can be further applied to many other implementations of meshfree methods, especially those implementations that require irregular
particle distributions, adding and deleting of particles and leverage adaptive background grids.
\end{abstract}

@@

SPH is a meshless scheme, initially developed for astrophysical applications by Lucy\cite{lucy1977numerical} and Gingold and Monaghan\cite{gingold1977smoothed}. 
Subsequently it was extended to computational fluid mechanics. 

@@

simple scenarios like breaking-waves, floodings etc. Work on more complicated problems, such as eruption of volcano plume (figure \ref{fig:Plume}) which is essentially a mult-phase turbulent flow,

@@

Prediction of such complicated phenomena with acceptable accuracy at given time window requires resolutions (very high particle counts) that cannot been accomplished without parallel computing. What's more, imposing of some types of boundary conditions (such as realistic wind field, eruption boundary condition) requires dynamically adding and removing of particles during simulation. This requires efficient and more flexible data management scheme. What we need to point out is that the benefit of flexible data management strategies are not only limited to specific implementations of SPH. Actually, flexibility in data management are more critical for several advanced techniques of SPH, such as dynamic particle splitting techniques\cite{vacondio2012accurate, feldman2007dynamic}, which will give greater resolution at the area of interest by splitting one large particle to several smaller ones.

@@

To the best of the author's knowledge, no implementation of SPH has the feature of adjusting computational domain based on simulation needs. For JPUE simulation, such feature will greatly reduce computational cost by avoiding computing of uninfluenced fluid. This feature is accomplished by adding a scan function to monitor the outermost layer of the domain and turn ghost particles to real particles at proper time.\\

@@

The data structure, particle and bucket indexing strategies, domain decomposition, dynamic load balancing method and domain adjusting strategies in this paper can be easily adopted by other implementations of any meshfree methods(include SPH). The flexibility of data accessing enables implementing of meshless methods for solving of more complicated problems and using of more advanced techniques.

@@

\caption{Basic work flow for SPH code}

@@

SPH is a meshfree, Lagrangian method. The domain is discretized by particles and the position of each particle is updated at every time step. The physical laws (such as conservation laws of mass, momentum and energy) written in the form of partial differential equations need to be transformed into the Lagrangian particle formalism of SPH. Using a kernel function that provides the weighted estimate of the field variables in the neighborhood of a discrete point (particle), the integral equations are evaluated as sums over neighbor particles. Only particles located within support of kernel function will interact. Thus, physical properties (density, velocity, internal energy...) are updated based on its neighbors. So a neighbor searching needs to be carried out before updating of physical properties. We use buckets which contain all particles associated with a sub-domain 

@@

SFC-based indexing scheme can guarantee uniqueness of particle identifier only in simple scenarios when particles are added once while setting up initial condition. In some situations, new particles need to be added while simulation. For example, new particles need to be added at the bottom of the eject vent for JPUE simulation. To distinguish particles added at the "same place" (the small area, all points within which will be mapped to the same $k$.) at different time steps, we extend the SFC-based key to time-dependent SFC based key by including date of birth of particles into the key. The time-dependent SFC based key can be written as: $[k,t]$, where $t$ is the time step. The map $h_n$ will become:

@@

We need to emphasize here that in theory motion of particles will destroy locality established based on initial coordinates of particles. However, as particles are moving fairly regularly in a JPUE, the locality of most of particles is largely conserved during simulation. We will revisit this question for more chaotic flows and 

@@

The most basic data structure of SPH are particle, for problem description, and bucket, for neighbors searches and domain decomposition. Both are defined as classes in C++. Information that is contained in particle class can be categorized into six categories: ID(the key), affiliate(rank of the process that the particle belongs to), primitive variables (variables show up as unknowns in governing equations, e.g. density, velocity), secondary variables (physical properties that can be computed from primitive variables, they are stored to avoid repeated computations eg. pressure, temperature), flags (indicators, such as indicator for ghost particle and real particles, indicator for particles of different phases) and neighbor information (it is a vector of particle keys in our application). Similarly,  Information that is contained in  a bucket class can also be categorized into different categories: ID, affiliate, domain information (maximum and minimum coordinates, boundary information), flags (indicators, such as indicator for guest and non-guest, indicator for active and inactive), neighbor information (keys of 27 neighbor buckets including its own key) and owned particles. Objects defined based on these two classes are then accessed through hash tables.

@@

One way to handle hash conflicts is by using an additional sorted vector attached to the hash table. When several keys hash to the same slot, a vector will be created. The vector is sorted based on keys so that a binary search can be used to find the correct position for adding, deleting or retrieving. Another option to handle hash conflicts is using an additional link list which is more flexible in memory allocation. The average time complexity of binary search is O(log n) while that for linear search based on link list is O(n). However, accessing efficiency of link list is much lower than array based data structure, especially when the link list becomes longer. Choosing of proper way to handle hash conflicts greatly depends on the problem itself. For the test problem in this paper, successively adding particles at the bottom of the eject vent will lead to hash conflicts of many particles, which implies a long link list. But considering the very long conflicts only occur on several slots among millions (see figure \ref{fig:Particle_adding_with_link}), we still choose the link list to handle hash conflicts. This decision was made based on numerical experiments.

@@

Particles used in the test problem can be categorized into four types based on the particle-type-flag: real particle, wall ghost particle, pressure ghost particle and eruption ghost particle. Ghost particles are for imposing of corresponding boundary conditions (see figure \ref{fig:Domain_adjusting}). As different types of particles involve different amount of computational work, shown by table \ref{tab:Computational_cost_particles} and table \ref{tab:Computational_cost_steps}, we assign different work load weight for different types of particles based on profilling data. Instead of simply using number of contained particles as work load for bucket, work load of each bucket is determined by summing up work load weight of all particles within the bucket. The SFC sequence passing through centroids of all buckets now becomes a weighted sequence. Domain decomposition  is conducted based on the weighted SFC of buckets.

@@

    Particles & NS & UPME & UPD & UPP &VF &WBC &Total\\

@@

    wall & No & No & No & No & No & Yes &0.75\\

@@

    eruption & No & No & No & Yes & No & No & 0.02\\

@@

    pressure & No & No & No & No & No & No & 0.00\\

@@

As some of the neighbor particles reside in other partitions. A set of  particles and buckets are used to synchronize data across partitions. To minimize communications, data is synchronized only where needed, using non-blocking MPI communications. 

@@

\includegraphics[scale=0.36]{Work_flow_adjust}

@@

@@

@@

The work flow with domain adjusting is in figure \ref{fig:Work_flow_adjust}.
\subfloat[Computational domain at $t=t0$]{\includegraphics[scale=0.23]{Involved_initial}}
\subfloat[Computational domain at $t=t0 + \Delta t$]{\includegraphics[scale = 0.23]{Involved_domain_adjusting}}

@@

In current model the initial domain is a 3D box. 
%The boundaries are categorized into eruption vent (a circle area at the center of the bottom), wall boundary (box bottom), pressure boundary (Other faces of the box).
At the eruption vent, temperature $T$, velocity $\textbf{v}$, pressure $p$, initial mass fraction of solid portition $n_{g0}$ and mass discharge rate $\dot M$ is given. 
%The radius of vent is determined from $\rho$, $\dot M$ and $\textbf{v}$.
Non-slip, adiabatic wall boundary condition is imposed on the wall. The pressure of the surrounding atmosphere is specified on the pressure boundaries. 
%Except for the pressure, density, velocity, and energy will depend on the solution. We adpot ghost particles to impose these different type of boundary conditions (see figure \ref{fig:Domain_adjusting}). 

@@

There are several review papers \cite{monaghan1992smoothed, monaghan2005smoothed, price2012smoothed, monaghan2012smoothed} which provide a pretty comprehensive descriptions of SPH. These may be consulted for additional detail.%We will not cover the basic theory of SPH in this paper.

@@

\textbf{v}_{\alpha \beta}=\textbf{v}_{\alpha}-\textbf{v}_{\beta}

@@

\dfrac{d \textbf{r}_a}{dt} = \textbf{v}_{\alpha} \label{eq:gov-update-pos}

@@

A typical simulation result of volcano plume is shown in figure \ref{fig:Plume}. 
\subfloat[A plume from Sakurajima Volcano eruption\cite{PlumePhoto}]{\includegraphics[width=3.8cm,height=4.5cm]{plume_photo}}
\label{fig:plume_photo}
\subfloat[Contour of mass fraction of erupted material at 290 seconds]{\includegraphics[width=3.8cm,height=4.5cm]{Plume_simulation}}
\label{fig:plume_simulation}
\caption{The volcano plume}
\label{fig:Plume}
